Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-08-26 paper/paper.tex
small additions to section 2 in paper draft

\section{Simplicity and expressivity in semantics, pragmatics, and beyond}
At least since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that languages are shaped by trade-offs of competing pressures has played a pivotal role in synchronic and diachronic analyses (e.g. \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}). A cornerstone principle, both in linguistics and cognitive science more generally, is that of simplicity \citep{chater+vitanyi:2003}.

@@

More particular to natural communication, languages are pressured towards expressivity and learnability. Their opposition becomes clear when considering their consequences in the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). On the one side, a language with a single form is extremely easy to learn but lacking in expressivity. On the other, a language that associates a distinct form with all possible meanings speakers may want to convey is expressive but challenging to acquire. However, acquiring a system to express a potentially infinite set of meanings through finite means, the so-called transmission bottleneck \citep{kirby:2002}, is not the only problem learners confront. Instead, the relevant challenge for our purpose is that of selecting between functionally similar, if not identical, lexical meanings.

\hl{explain scalar implicatures and the challenge that arises, then go for simplicity in lexical entries being codified as a preference in the learners}

@@

%While his rationalization is not in want of detractors (see \citealt{piantadosi:2014} for a recent overview), 
\begin{itemize}
  \item \hl{Idea in LOT (\citealt{piantadosi+etal:underreview})}
\end{itemize}

@@

The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g. {not all} as hypothetical `nall') are unattested across languages has received substantial support (most prominently in \citealt[252-267]{horn:1984} but also e.g. in \citealt{horn:1972,traugott:2004,vdAuwera:2010}). To the best of our knowledge, this claim stands unchallenged.

>>>>

\section{Simplicity and expressivity in semantics, pragmatics, and beyond}
At least since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that languages are shaped by trade-offs of competing pressures has played a pivotal role in synchronic and diachronic analyses (e.g. \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}). A cornerstone principle, both in linguistics and cognitive science more generally, is that of simplicity \citep{chater+vitanyi:2003}. \hl{expand a bit here to prepare for linkage further down}

@@

More particular to natural communication, languages are pressured towards expressivity and learnability. Their opposition becomes particularly clear when considering their consequences in the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). On the one side, a language with a single form is extremely easy to learn but lacking in expressivity. On the other, a language that associates a distinct form with all possible meanings speakers may want to convey is expressive but challenging to acquire. However, acquiring a system to express a potentially infinite set of meanings through finite means, the so-called transmission bottleneck \citep{kirby:2002}, is not the only problem learners confront. Instead, for our purpose the relevant challenge is that of selecting between functionally similar, if not identical, pragmatically enriched lexical meanings.

A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf. \citealt{horn:1972,gazdar:1979}). These inferences are licensed for groups of expressions ordered in terms of informativity, here understood an entailment induced order.\footnote{\hl{maybe discuss that there are other possible orders considering relevant alternatives}}. For instance, {\em some} is entailed by {\em all} -- if it were true that `All students came to class', it would also be true that `Some students came to class'. However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey. Instead, the use of a less informative expression when a more informative one could have been used can license the inference that the more stronger alternative does not hold. That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold. In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class' through a pragmatically supplied upper-bound that rules out stronger alternatives. Analogously, a speaker can rely on her interlocutor to draw this inference without having to express the bound overtly, e.g. by stating {\em some but not all}.

Rephrasing our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives regularly selected for over the alternative of incorporating the bound in the semantics? This question is particularly striking when considering the diversity of the classes of expressions that license such inferences across languages (\citealt{horn:1972},\citealt[252-267]{horn:1984},\citealt{traugott:2004,vdAuwera:2010}). In the following, we assume learners to be biased, a priori, towards simpler (more compressed) representations, corresponding to the argument that rational learners should prefer simpler over more complex explanations of data \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.

\hl{summarize assumptions and then move into the dynamics right away}

%The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g. {not all} as hypothetical `nall') are unattested across languages has received substantial support (most prominently in \citealt[252-267]{horn:1984} but also e.g. in \citealt{horn:1972,traugott:2004,vdAuwera:2010}). To the best of our knowledge, this claim stands unchallenged.

@@

%While his rationalization is not in want of detractors (see \citealt{piantadosi:2014} for a recent overview),

@@
